{\plant}: A package for modelling forest trait ecology \& evolution

### 2015-05-29 ms.tex
Plot averages with respect to height across metapopulation.
For #13

Squashed commit of the following:

commit 1d0640e721e018abc5ba7dfc459a45810ebe6172
Author: Daniel Falster <daniel.falster@mq.edu.au>
Date:   Thu May 28 20:34:40 2015 +1000

    Increase resolution

commit d8df1369e081494d722500912f395b5e02079c65
Author: Daniel Falster <daniel.falster@mq.edu.au>
Date:   Thu May 28 18:56:47 2015 +1000

    interpolate using spline_loglog

commit 85cec05505f81e0f17f692dd60d5fc98b659b9e2
Author: Daniel Falster <daniel.falster@mq.edu.au>
Date:   Thu May 28 17:09:50 2015 +1000

    First cut at get

\documentclass[a4paper,11pt]{article}
\usepackage[osf]{mathpazo}
\usepackage{ms}
\usepackage[sort&compress]{natbib}
\raggedright

@@

\begin{document}
\mstitlepage
\noindent
\parindent=1.5em
\addtolength{\parskip}{.3em}
\doublespacing
\linenumbers
\section{Summary}\label{abstract}
\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\itemsep1pt\parskip0pt\parsep0pt
\item
  Population dynamics in forests are strongly size-structured:
  larger plants shade smaller plants while also expending
  proportionately more energy on building and maintaining woody stems.
  Although the importance of size-
  structure for demography is widely recognised, many mechanistic models
  either omit it entirely, or include only coarse approximations.
\item
  Here we introduce the {\plant} package (TRait Ecology and Evolution), an
  extensible framework for modelling plant demography across the entire
  life cycle and in structured metapopulations, via coupled differential equations.
  At its core, {\plant} is an
  individual-based mechanistic model where plant physiology and demography is mediated by
  traits. Individuals from multiple species can be grown in isolation,
  in patches of competing plants, or in metapopulations under a
  disturbance regime. Dynamics within patches of competing plants are
  resolved using novel extensions of the ``Escalator Boxcar Train''
  technique. Combined effects of trait-, size- and patch-structured
  dynamics are integrated into population level estimates of
  reproductive fitness.
\item
  {\plant} is an open source R package and is available at
  \href{https://github.com/traitecoevo/plant}{github.com/traitecoevo/plant}.
  While accessed from R, the core routines in {\plant} are written in C++.
  The package provides for alternative physiologies and for
  hyper-parameterisation to capture trade-offs among parameters. A
  detailed test suite is provided to ensure correct behaviour of the code.
\item
  We provide worked examples illustrating {\plant}'s use to study the
  influence of functional traits on growth of individual plants, of
  whole patches and on assembly of ecological communities.
\end{enumerate}

@@

Plant growth and demography are fundamentally size- and trait-
structured, influencing dynamics over time-scales ranging from
instantaneous physiological effects to long-term evolutionary outcomes
\citep{Harper-1977, Westoby-2002}. As an individual increases its leaf
area, it increases its potential to generate photosynthate.  In
contrast, as indivuduals grow, they must use increasing fractions of
this photosynthetic income to building and maintaining support tissues
rather than to new leaves \citep{Givnish-1988, Enquist-2007} and
reproduction \citep{Thomas-2011}. Consequently rates of growth and
mortality change with individual size \citep{Muller-2006, Thomas-2011,
  Ruger-2011, Wright-2010}.
%
The patterns of changing growth rate vary by plant traits; leaf
construction costs and wood density affect growth rates and rates of
tissue replacement \citep{Wright-2010}, while seed size and height at
maturation determine start and end points of ontogenetic trajectories
\citep{Westoby-2002}
%
Strong feedbacks emerge between individuals within a forest due to
competition for light and other resources, such that the growth rate
of one individual depends on the size and traits of nearby
individuals.  This makes it difficult to predict demographic processes
without modelling both individual growth and the interactions between
individuals.
% TODO[RGF]: link back things about natural selection, successional
% dynamics, self thinning, trait evolution and species coexistance.

% NOTE[RGF]: I thought this was way overrefrenced and so just kept a
% few oldies.  Please re-add any key ones that you want to see here.
Although the importance of size-structure for vegetation function and
demography has been long been recognised
\citep[e.g.,][]{Harper-1977, Shugart-1980, Hara-1984},
most large-scale mechanistic models of plant growth either omit it
entirely, or only include coarse approximations
\citep{Cramer-2001, Dekauwe-2014, Kelley-2013, Sitch-2003}. The likely
reason is the computational
challenge of modelling size-structured dynamics. 
% TODO[RGF]: I don't see the link with the above?  Can this be
% shortened/tightened?  It's too long for a 3k word paper
During the 1980s and
early 1990s, size-structured ``gap'' models were widely used
\citep[e.g.][]{Huston-1987, Shugart-1980}. These gap models were
parameterised in terms of growth using statistical functions; this meant 
that data had to be collected for each species, limiting widespread application. 
The next generation of models focussed on establishing core physiology for 
a limited number of functional types, and
sacrificed details of demography in order to achieve global coverage
\citep{Haxeltine-1996,Woodward-1995}. This history has brought us to the
current situation where all but a couple
\citep[e.g.][]{Moorcroft-2001, Smith-2014} of the most popular vegetation
models lack size-structure. Instead, most models
are focused on modelling fluxes carbon and water, but say little about
demographic behaviour, and consequently are also unable to account for
any non-linear demographic feedbacks that may arise.

Similarly, most models used in theoretical ecology to explore
questions about species coexistence also lack size-structured dynamics
\citep[e.g.][]{Calcagno-2006, Geritz-1995, Leimar-2013,
  Levin-1974,MacArthur-1967,Tilman-1985}.  It is common to assume
competitive interactions are influenced by size-related traits such as
adult size and seed size, but detailed size-structured demography is
rarely considered, at least for plant models \cite[for animal
examples, see][]{Deroos-1988, Deroos-1992}. Consequently, it has been
difficult to connect mechanisms described in these simplistic models
with type of data often collected in the field, such as physiological
leaf traits, and demographic data on growth and mortality in long-term
monitoring plots \citep{Adler-2013}.

In this note, we introduce the {\plant} package for R \citep{R-2015};
a mechanistic framework for studying the effects of size structure and
trait variation on the demography of individual plants, of patches of
competing plants, and of meta-populations structured by a prevailing
disturbance regime.
%
The {\plant} package builds on much existing work investigating the
effect of traits on vegetation properties and the mechanisms
facilitating coexistence of trait mixtures
\citep{Kohyama-1993,Deroos-1997,Moorcroft-2001,Falster-2011,Falster-2015}.

% TODO[RGF]: restructure here - I'm getting lost
{\plant} provides a platform for investigating how morphological and
physiological traits influence growth rate and shade tolerance, thus
potentially giving a theoretical foundation for observed empirical
patterns \citep{Wright-2010,Baltzer-2007}.

% NOTE[RGF]: I dropped the appeal to novelty (growing an individual
% plant) because I don't find them very convincing.
While the applications highlighted reflect our own interests in understanding
trait-based demography and community assembly, we expect the methods provided
to be useful in variety of contexts. . Being driven by traits,
the model can be extended to potentially very many species. Users can also 
add their own physiological models. Individual plants
can then be  situated in a patch with competitors and the demography of a
competing population  investigated, thereby capturing phenomena such as multi-
species self-thinning,  successional transitions and age-related changes in
vegetation productivity. {\plant}  models the emergent behaviour of
patches deterministically, based on a continual  size-density distribution. As
such, it is not required to specify a particular patch size.  Finally,
{\plant} allows for emergent phenomena across a metapopulation of patches
to  be investigated, and thus provides a tractable mechanism to scale from
individual plant to  vegetation, incorporating the non-linear effects of size
and traits on vegetation outcomes. In  this sense {\plant} is similar to
the Ecosystem Demography \citep{Moorcroft-2001}  and LPJ-GUESS
\citep{Smith-2014} models. However, whereas these other models were designed
to primarily capture large-scale regional biogeocehcmisty driven by climatic
variables,  {\plant} was designed so that the rules driving the
demography of individual trees, patches and metapopulations could be
manipulated and studied in detail. The three models are therefore highly 
complimentary, and also connect with the increasingly sophisticated statistical 
methods now available to analyse size-structured data
\citep[e.g.][]{Metcalf-2013}.

We describe the general approach of the package and then sections with
details at different levels of ecological organisation. Each section
provides a short description of the required methods, design
considerations, and then example applications. Detailed technical
documents are provided as supplementary information (see Appendices),
and updated versions of technical documents are also available within
the {\plant} package itself.

@@

% NOTE: Keep the repetition with above under control.
% NOTE: LV are not really mechanistic

{\plant} is a mechanistic model, meaning the dynamics of the system
arise from rules specifying how individuals grow and interact.  The
core rules in {\plant} are about the short-term physiological
functioning of an individual plant and how these are influenced by its
traits, size and light environment.
%
On top of this physiolgical model, the package implements methods for
population and adaptive dynamics (Fig.  \ref{fig:schematic}) following
methods described by \citet{Falster-2011} and
\citet{Falster-2015}. 
%
The dynamics at higher levels of organisation arise as emergent
properties, driven by growth physiology, competition for light and
disturbance (Fig.  \ref{fig:schematic}). Demographic phenomena can be
studied at the three levels: individual plants, stands of competing
plants, and entire metapopulations.
%
We developed {\plant} as an extended thought experiment for helping to
understand competition and succession by having these driven by
low-level processes rather than being explicitly implemented in the
model.
% NOTE: I think that foreshaddowing some of the competition kernels
% idea here is sensible.
%
% TODO: Somewhere, perhaps above, perhaps highlight the issue of doing
% coexistance modelling with explicit modelling - the models are so
% well understood it's easy to drive the behaviour.  But we want to
% have something where we just get coexistence to study how tradeoffs
% generate it.
%
% TODO: perhaps salavageable?
%
%   {\plant} was developed with this style of analysis in mind, while
%   also allowing for a richer set of ecological dynamics than is
%   possible in population models lacking size structure.

@@

The core of {\plant} is a model for an individual plant's
physiological strategy as specified by its traits
(Fig. \ref{fig:schematic}a). This model estimates rate of biomass
production for a plant, given its size and a light environment. Gross
photosynthetic income is estimated from total leaf area and the light
distribution across the plant's canopy. Costs of tissue respiration
and turnover are then subtracted. The remaining biomass is then
allocated between growth and reproduction.  
%
The key outputs needed by higher levels (see below) is growth rate,
mortality rate and rate of seed production.  However additional
quantities are also computed, such as total assimilation and the
respiration, turnover, and allocation rates for different tissues (see
Appendix \ref{sec:plant})

The default physiological model used in {\plant} largely reflects that
presented in \citet{Falster-2011}, but with extensions allowing for
diameter growth also to be estimated. To properly model trait
variation, different parameters (such as rates of leaf turnover,
respiration, photosynthesis per leaf area) in the model are linked via
assumed tradeoffs. For example, the trait LMA is used to estimate the
rate of leaf turnover, based on observed scaling relationship spanning
across diverse vegetation and plant functional global
\citep{Wright-2004} (see Appendix \ref{sec:FFW16} for details).

The current parametrisation of the physiological model is based on
\citet{Falster-2011}. Parameter values were drawn from available
literature (see Appendix \ref{sec:FFW16}, table 3); however for some
parameters limited or no data were readily available. Providing
updated parameter sets is trivial; the challenge is to infer suitable
values by constraining outputs from {\plant} against diverse data
sources \citep{Lebauer-2012, Keenan-2013}. This is an area in which we
are still actively working, and one of the major use cases we envisage
for {\plant}.

\subsection{Design features}

While we have implemented one physiological model (\texttt{FFW16}),
the package is designed to allow arbitrary additional physiologies to
be introduced.  That is, rather than varying the parameters of the
equations, the equations themselves can be changed, additional
resources could be competed for, and the higher level functionality of
the package can still be used.  As an intermediate level, we allow for
parameters to imply relationships among arbitrary other parameters,
allowing straightforward modelling of trade-offs.  For example,
setting leaf mass per unit area (LMA) affects ....
% TODO[RGF]: please complete (running out of time here)

@@

With the basic physiological model, {\plant} can be used to estimate
essential physiological rates for individual plants
(Fig. \ref{fig:plant}). The function \texttt{grow\_plant\_to\_size}
takes a given strategy and light environment and grows the plant,
producing a trajectory of size over time (Fig. \ref{fig:plant}a). This
is achieved by integrating an Ordinary Differential Equation (ODE) of
size-dependent growth rate (see Appendices \ref{sec:demography} and
\ref{sec:plant} for more details). As the plant grows, allocation to
different tissue types varies, and the composition of the plant
changes to include more stem and less leaf
(Fig. \ref{fig:plant}b). This in turn affects maintenance and turnover
costs and the growth rate of plants (Fig. \ref{fig:plant}c).  This
shift in allocation alone generates a distinctive hump-shaped pattern
of absolute growth with size \citep{King-2011}, and equally the widely
regarded decline in relative growth rate with size from birth onwards
\citep{Enquist-2007}.

By varying the light environment and measuring growth rate, the whole
plant light compensation point (WPLCP) can be computed (Fig.
\ref{fig:plant}d). The WPLCP is the light level at which the plant 
stops growing, and is increasingly regarded as the most useful
measure of a strategy's shade tolerance
\citep{Givnish-1988, Baltzer-2007, Lusk-2013}. As expected, WPLCP
increases with plant size, due to increased costs of building and
maintaining stem and leaf tissues \citep{Givnish-1988}. Likewise, WPLCP
decreases with LMA, because high LMA species have lower leaf turnover
\citep{Baltzer-2007, Lusk-2013}.

@@

% TODO: missing something saying we are considering an infinitely
% large population here.
Modelling the development of a competitive hierarchy within a patch
requires that both the initial size distribution and the inflow of new
recruits be specified. We have primarily been interested in dynamics of
a patch recovering from disturbance, and have therefore started with an
empty patch and constant flow of seeds from a global seed rain (Fig.
\ref{fig:schematic}b).

% TODO: This is not really a design feature. That heading is almost
% extinct in the plant section and missing in the last - perhaps just
% merge with the above section?
\subsection{Design features}

When modelling a patch, we are interested in modelling changes in the
size-density-distribution \(n(x,h,a)\) with patch age \(a\), i.e. the density of
individuals with traits \(x\) and height \(h\) within a patch of age \(a\). Density
is measured as individuals per unit height and per unit ground area.

We assume that patches are large, and are vertically but not horizontally
structured. In other words, we account for size differences but not spatial 
layout within the patch. Similar assumptions are found in many models
simulating size-structured dynamics
\citep{Moorcroft-2001, Huston-1987, Smith-2014,Kohyama-1993}. The assumption of a large
patch means the population of plants within the patch -- and thus the 
dynamics of \(n\) -- behave deterministically \citep{Deroos-1997}, and can therefore be
approximated via a Partial Differential Equation (PDE) (see Appendix
\ref{sec:demography}). The same PDE captures average behaviour
across a large number of small patches \citep{Moorcroft-2001}. The model may 
therefore capture average demographic behaviour across a range of patch sizes.

Ideally one would also consider spatial interactions within patches,
however such models are very computationally demanding, and also
impossible to solve in a deterministic fashion \citep{Huston-1987, 
Pacala-1996}.
% TODO: I added "useful" in here, but more changes perhaps?
Moreover, it remains unclear whether resolving spatial details within a
patch would provide further useful ecological detail, in part because the
process of competitive thinning tends to break down spatial clusters
\citep{Strigul-2008}.

Our approach for modelling size-structured population dynamics
is based on the Escalator Boxcar Train technique (EBT)
\citep{Deroos-1997, Deroos-1992, Deroos-1988}. The EBT solves the PDE
describing development of \(n\) by approximating the density function
with a collection of cohorts spanning the size spectrum. Following a
disturbance, a series of cohorts are introduced into each patch. These
cohorts are then transported along the characteristic equations of the
PDE. Biologically these are the growth trajectories of individuals,
conditioned on them surviving (e.g. Fig. \ref{fig:patch}a). Meanwhile,
growth and mortality combine to alter the density of individuals along
the trajectories (Fig. \ref{fig:patch}b).

% TODO: early in this paragraph include a pointer to the appropriate
% vignette that justifies it
We introduce two extensions to the EBT: i) a new method for estimating
the size-density distribution, and ii) a technique for handling strong
size-asymmetric competitive feedbacks, such as occur under strong
competition for light. The original EBT \citep{Deroos-1997,
  Deroos-1992, Deroos-1988} proceeds by approximating the first and
second moments of the density distribution \(n\) within each
cohort. In {\plant} we use a different approach. Instead of tracking
first and second moments of the size-distribution within cohorts, we
track a point mass estimate of \(n\) along the trajectory for the
boundary of each cohort. We found this approach preferable because it
does not require us to maintain a separate set of equations for the
smallest cohort \citep{Deroos-1997}.

The extension for handling strong size-asymmetric competition involves
adaptive refining of time points at which new cohorts are introduced
into the population. In the original EBT, cohorts were introduced at a
fixed time interval. Under competition, the growth trajectories of
individuals born at similar times can diverge substantially over time
(Fig. \ref{fig:patch}a; Appendix \ref{sec:demography},
\ref{sec:patch}, and \ref{sec:cohort-spacing}). This is can become a
problem later during patch development because much of the population
can be located between widely-spaced cohorts, delivering low numerical
precision. To maintain numerical precision we therefore use an
iterative algorithm to adaptively refine the cohort spacing until the
trajectories of adjacent cohorts remain adequately resolved throughout
development of the patch (see Appendix \ref{sec:demography} for
details). This refinement causes an uneven distribution of cohort
introduction times across patch age (Fig.  \ref{fig:patch}a), with
considerably tighter spacing of cohorts in younger patches.

@@

% TODO: not a paragraph.  Delete?  Or merge together with one of the
% others?
Modelling of size-structured dynamics via solving of a deterministic PDE
enables multiple complex non-linear demographic phenomena to be
investigated.

First, {\plant} can simulate patterns of patch development after
disturbance, including self-thinning behaviours and successional
replacement. Whereas previous approaches for modelling self-thinning
were limited to a single species
\citep[e.g.][]{Barnes-2004, Coomes-2007}, {\plant} accommodates multiple
 species with different traits (Fig. \ref{fig:schematic}b; Fig.
\ref{fig:patch}).

>>>>

\documentclass[a4paper,11pt]{article}
\usepackage[osf]{mathpazo}
\usepackage{ms}
\usepackage[]{natbib}
\raggedright

@@

\begin{document}
\mstitlepage
\noindent
\parindent=1.5em
\addtolength{\parskip}{.3em}
\doublespacing
\linenumbers
\section{Summary}\label{abstract}
\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\itemsep1pt\parskip0pt\parsep0pt
\item
  Population dynamics in forests are strongly size-structured:
  larger plants shade smaller plants while also expending
  proportionately more energy on building and maintaining woody stems.
  Although the importance of size-
  structure for demography is widely recognised, many mechanistic models
  either omit it entirely, or include only coarse approximations.
\item
  Here we introduce the {\plant} package, an
  extensible framework for modelling trait-based demography, ecology and evolution 
  in vegetation, via coupled differential equations.
  At its core, {\plant} is an
  individual-based mechanistic model where plant physiology and demography is mediated by
  traits. Individuals from multiple species can be grown in isolation,
  in patches of competing plants, or in metapopulations under a
  disturbance regime. Dynamics within patches of competing plants are
  resolved using novel extensions of the ``Escalator Boxcar Train''
  technique. Combined effects of trait-, size- and patch-structured
  dynamics are integrated into population level estimates of
  reproductive fitness.
\item
  {\plant} is an open source R package and is available at
  \href{https://github.com/traitecoevo/plant}{github.com/traitecoevo/plant}.
  While accessed from R, the core routines in {\plant} are written in C++.
  The package provides for alternative physiologies and for
  hyper-parameterisation to capture trade-offs among parameters. A
  detailed test suite is provided to ensure correct behaviour of the code.
\item
  We provide worked examples illustrating {\plant}'s use to study the
  influence of functional traits on growth of individual plants, of
  whole patches and on assembly of ecological communities.
\end{enumerate}

@@

Plant growth and demography are fundamentally size- and trait-
structured, influencing dynamics over time-scales ranging from
instantaneous physiological effects to long-term evolutionary outcomes
\citep{Harper-1977, Westoby-2002}. As an individual increases its leaf
area, it increases its potential to generate photosynthate.  In
contrast, as individuals grow, they must use increasing fractions of
this photosynthetic income to building and maintaining support tissues
rather than to new leaves \citep{Givnish-1988, Enquist-2007} and
reproduction \citep{Thomas-2011}. Consequently rates of growth and
mortality change with individual size \citep{Muller-2006, Ruger-2011, Thomas-2011}.
%
The patterns of changing growth rate vary by plant traits; leaf
construction costs and wood density affect growth rates and rates of
tissue replacement \citep{Wright-2010}, while seed size and height at
maturation determine start and end points of ontogenetic trajectories
\citep{Westoby-2002}
%
Strong feedbacks emerge between individuals within a forest due to
competition for light and other resources, such that the growth rate
of one individual depends on the size and traits of nearby
individuals.  This makes it difficult to predict demographic processes
without modelling both individual growth and the interactions between
individuals.
% TODO[RGF]: link back things about natural selection, successional
% dynamics, self thinning, trait evolution and species coexistance.

Although the importance of size-structure for vegetation function and
demography has been long been recognised
\citep[e.g.,][]{Harper-1977, Shugart-1980, Huston-1987},
most large-scale mechanistic models of plant growth either omit it
entirely, or only include coarse approximations
\citep{Cramer-2001, Sitch-2003, Kelley-2013,Dekauwe-2014}. The likely
reason is the computational
challenge of modelling size-structured dynamics. 
% TODO[RGF]: I don't see the link with the above?  Can this be
% shortened/tightened?  It's too long for a 3k word paper
During the 1980s and
early 1990s, size-structured ``gap'' models were widely used
\citep[e.g.][]{Shugart-1980,Huston-1987}. These gap models were
parameterised in terms of growth using statistical functions; this meant 
that data had to be collected for each species, limiting widespread application. 
The next generation of models focussed on establishing core physiology for 
a limited number of functional types and
sacrificed details of demography in order to achieve global coverage
\citep{Woodward-1995,Haxeltine-1996}. This history has brought us to the
current situation where all but a couple
\citep[e.g.][]{Moorcroft-2001, Smith-2014} of the most popular vegetation
models lack size-structure. Instead, most models
are focused on modelling fluxes carbon and water, but say little about
demographic behaviour; consequently they are also unable to account for
any non-linear demographic feedbacks that may arise.
% While many vegetation models still lack size structure, the importance of
% size-structure for modelling phenomena such as allocation and production is
% being increasingly noted \citep{Moorcroft-2001, Dekauwe-2014, Falster-2011}.

Similarly, most models used in theoretical ecology to explore
questions about species coexistence also lack size-structured dynamics
\citep[e.g.][]{MacArthur-1967,Levin-1974,Tilman-1985,Geritz-1995, 
Calcagno-2006, Leimar-2013}.  It is common to assume
competitive interactions are influenced by size-related traits such as
adult size and seed size, but detailed size-structured demography is
rarely considered, at least for plant models \cite[for animal
examples, see][]{Deroos-1988, Deroos-1992}. Consequently, it has been
difficult to connect mechanisms described in these simplistic models
with type of data often collected in the field, such as physiological
leaf traits and demographic data on growth and mortality in long-term
monitoring plots \citep{Adler-2013}.

In this note, we introduce the {\plant} package for R \citep{R-2015};
a mechanistic framework for studying the effects of size structure and
trait variation on the demography of individual plants, of patches of
competing plants and of meta-populations structured by a prevailing
disturbance regime.
%
The {\plant} package builds on much existing work investigating the
effect of traits on vegetation properties and the mechanisms
facilitating coexistence of trait mixtures
\citep{Kohyama-1993,Deroos-1997,Moorcroft-2001,Falster-2011,Falster-2015}.

% TODO[RGF]: restructure here - I'm getting lost
{\plant} provides a platform for investigating how morphological and
physiological traits influence growth rate and shade tolerance, thus
potentially giving a theoretical foundation for observed empirical
patterns \citep{Baltzer-2007,Wright-2010}.

% NOTE[RGF]: I dropped the appeal to novelty (growing an individual
% plant) because I don't find them very convincing.
While the applications highlighted reflect our own interests in understanding
trait-based demography and community assembly, we expect the methods provided
to be useful in variety of contexts. . Being driven by traits,
the model can be extended to potentially very many species. Users can also 
add their own physiological models. Individual plants
can then be  situated in a patch with competitors and the demography of a
competing population  investigated, thereby capturing phenomena such as multi-
species self-thinning,  successional transitions and age-related changes in
vegetation productivity. {\plant}  models the emergent behaviour of
patches deterministically, based on a continual  size-density distribution. As
such, it is not required to specify a particular patch size.  Finally,
{\plant} allows for emergent phenomena across a metapopulation of patches
to  be investigated, thereby providing a workable mechanism to scale from
individual plant to  vegetation, incorporating the non-linear effects of size
and traits on vegetation outcomes. 
% To strigul?
In  this sense {\plant} is similar to
the \textsc{Ecosystem Demography} \citep{Moorcroft-2001}  and \textsc{LPJ-GUESS}
\citep{Smith-2014} models. However, whereas these other models were designed
to primarily capture large-scale regional biogeochemistry driven by climatic
variables,  {\plant} was designed so that the rules driving the
demography of individual trees, patches and metapopulations could be
manipulated and studied in detail. The three models are therefore highly 
complimentary, they also connect with the increasingly sophisticated statistical 
methods now available to analyse size-structured data
\citep[e.g.][]{Metcalf-2013}.

We describe the general approach of the package and then sections with
details at different levels of ecological organisation. Each section
provides a short description of the required methods and then example 
applications. Detailed technical
documents are provided as supplementary information (see Appendices),
and updated versions of technical documents are also available within
the {\plant} package itself.

@@

% NOTE: Keep the repetition with above under control.

{\plant} is a mechanistic model, meaning the dynamics of the system
arise from rules specifying how individuals grow and interact.  The
core rules in {\plant} are about the short-term physiological
functioning of an individual plant and how these are influenced by its
traits, size and light environment.
%
On top of this physiological model, the package implements methods for
population and adaptive dynamics (Fig.  \ref{fig:schematic}) following
methods described by \citet{Falster-2011} and
\citet{Falster-2015}. 
%
The dynamics at higher levels of organisation arise as emergent
properties, driven by growth physiology, competition for light and
disturbance (Fig.  \ref{fig:schematic}). Demographic phenomena can be
studied at the three levels: individual plants, stands of competing
plants and entire metapopulations.
%
We developed {\plant} as an extended thought experiment for helping to
understand competition and succession by having these driven by
low-level processes rather than being explicitly implemented in the
model.
% NOTE: I think that foreshaddowing some of the competition kernels
% idea here is sensible.
%
% TODO: Somewhere, perhaps above, perhaps highlight the issue of doing
% coexistance modelling with explicit modelling - the models are so
% well understood it's easy to drive the behaviour.  But we want to
% have something where we just get coexistence to study how tradeoffs
% generate it.
%
% TODO: perhaps salavageable?
%
%   {\plant} was developed with this style of analysis in mind, while
%   also allowing for a richer set of ecological dynamics than is
%   possible in population models lacking size structure.

@@

The core of {\plant} is a model for an individual plant's
physiological strategy as specified by its traits
(Fig. \ref{fig:schematic}a). This model estimates rate of biomass
production for a plant, given its size and a light environment. Gross
photosynthetic income is estimated from total leaf area and the light
distribution across the plant's canopy. Costs of tissue respiration
and turnover are then subtracted. The remaining biomass is then
allocated between growth and reproduction.  
%
The key outputs needed by higher levels (see below) is growth rate,
mortality rate and rate of seed production.  However additional
quantities are also computed, such as total assimilation and the
respiration, turnover and allocation rates for different tissues (see
Appendix \ref{sec:plant})

The default physiological model used in {\plant} largely reflects that
presented in \citet{Falster-2011}, but with extensions allowing for
diameter growth also to be estimated. To properly model trait
variation, different parameters (such as rates of leaf turnover,
respiration, photosynthesis per leaf area) in the model are linked via
assumed tradeoffs. For example, the trait \textsc{lma} is used to estimate the
rate of leaf turnover, based on observed scaling relationship spanning
across diverse vegetation and plant functional global
\citep{Wright-2004} (see Appendix \ref{sec:FFW16} for details).

The current parametrisation of the physiological model is based on
\citet{Falster-2011}. Parameter values were drawn from available
literature (see Appendix \ref{sec:FFW16}, table 3); however for some
parameters limited or no data were readily available. Providing
updated parameter sets is trivial; the challenge is to infer suitable
values by constraining outputs from {\plant} against diverse data
sources \citep{Lebauer-2012, Keenan-2013}. This is an area in which we
are still actively working and one of the major use cases we envisage
for {\plant}.

While we have implemented one physiological model (\texttt{FFW16}),
the package is designed to allow arbitrary additional physiologies to
be introduced.  That is, rather than varying the parameters of the
equations, the equations themselves can be changed, additional
resources could be competed for and the higher level functionality of
the package can still be used.  As an intermediate level, we allow for
parameters to imply relationships among arbitrary other parameters,
allowing straightforward modelling of trade-offs.  For example,
setting leaf mass per unit area (\textsc{lma}) affects leaf turnover rate,
because it is known that leaves which are cheaper to build have shorter 
lifespan \citep{Wright-2004}.

@@

With the basic physiological model, {\plant} can be used to estimate
essential physiological rates for individual plants
(Fig. \ref{fig:plant}). The function \texttt{grow\_plant\_to\_size}
takes a given strategy and light environment and grows the plant,
producing a trajectory of size over time (Fig. \ref{fig:plant}a). This
is achieved by integrating an Ordinary Differential Equation of
size-dependent growth rate (see Appendices \ref{sec:demography} and
\ref{sec:plant} for more details). As the plant grows, allocation to
different tissue types varies and the composition of the plant
changes to include more stem and less leaf
(Fig. \ref{fig:plant}b). This in turn affects maintenance and turnover
costs and the growth rate of plants (Fig. \ref{fig:plant}c).  This
shift in allocation alone generates a distinctive hump-shaped pattern
of absolute growth with size \citep{King-2011} and equally the widely
regarded decline in relative growth rate with size from birth onwards
\citep{Enquist-2007}.

By varying the light environment and measuring growth rate, the whole
plant light compensation point (\textsc{wplcp}) can be computed (Fig.
\ref{fig:plant}d). The \textsc{wplcp} is the light level at which the plant 
stops growing and is increasingly regarded as the most useful
measure of a strategy's shade tolerance
\citep{Givnish-1988, Baltzer-2007, Lusk-2013}. As expected, \textsc{wplcp}
increases with plant size, due to increased costs of building and
maintaining stem and leaf tissues \citep{Givnish-1988}. Likewise, \textsc{wplcp}
decreases with \textsc{lma}, because high \textsc{lma} species have lower leaf turnover
\citep{Baltzer-2007, Lusk-2013} (Fig.
\ref{fig:plant}d).

@@

We assume that patches are vertically but not horizontally
structured. In other words, we account for size differences but not spatial 
layout within the patch. Similar assumptions are found in many models
simulating size-structured dynamics
\citep{Kohyama-1993, Huston-1987, Moorcroft-2001, Smith-2014}. In addition, 
we assume that patches are very large -- this means the population dynamics 
within the patch behave deterministically \citep{Deroos-1997} and can therefore 
be approximated via a Partial Differential Equation (\textsc{pde}) (see Appendix
\ref{sec:demography}). The same \textsc{pde} captures average behaviour
across a large number of small patches \citep{Moorcroft-2001}. The model may 
therefore capture average demographic behaviour across a range of patch sizes.

Ideally one would also consider spatial interactions within patches,
however such models are very computationally demanding and also
impossible to solve in a deterministic fashion \citep{Huston-1987, 
Pacala-1996}. Moreover, it remains unclear whether resolving spatial details within a
patch would provide further useful ecological detail, in part because the
process of competitive thinning tends to break down spatial clusters
\citep{Strigul-2008}.

When modelling a patch, we are interested in modelling changes in the size-
density-distribution \(n(x,h,a)\) with patch age \(a\), i.e. the density of
individuals with traits \(x\) and height \(h\) within a patch of age \(a\).
Density is measured as individuals per unit height and per unit ground area.
Modelling the development of  \(n(x,h,a)\) requires that both the initial size
distribution and the inflow of new recruits be specified. We have primarily
been interested in dynamics of a patch recovering from disturbance and have
therefore started with an empty patch and constant flow of seeds from a global
seed rain (Fig. \ref{fig:schematic}b).

Our approach for modelling size-structured population dynamics
is based on the Escalator Boxcar Train technique (\textsc{EBT})
\citep{Deroos-1988, Deroos-1992, Deroos-1997}. The \textsc{EBT} solves the \textsc{pde}
describing development of \(n\) by approximating the density function
with a collection of cohorts spanning the size spectrum. Following a
disturbance, a series of cohorts are introduced into each patch. These
cohorts are then transported along the characteristic equations of the
\textsc{pde}. Biologically these are the growth trajectories of individuals,
conditioned on them surviving (e.g. Fig. \ref{fig:patch}a). Meanwhile,
growth and mortality combine to alter the density of individuals along
the trajectories (Fig. \ref{fig:patch}b).

We introduce two extensions to the \textsc{EBT}: i) a new method for estimating
the size-density distribution and ii) a technique for handling strong
size-asymmetric competitive feedbacks, such as occur under strong
competition for light (see Appendix \ref{sec:demography} for
details). The original \textsc{EBT} \citep{Deroos-1988,
  Deroos-1992, Deroos-1997} proceeds by approximating the first and
second moments of the density distribution \(n\) within each
cohort. In {\plant} we use a different approach. Instead of tracking
first and second moments of the size-distribution within cohorts, we
track a point mass estimate of \(n\) along the trajectory for the
boundary of each cohort. We found this approach preferable because it
does not require us to maintain a separate set of equations for the
smallest cohort \citep{Deroos-1997}.

The extension for handling strong size-asymmetric competition involves
adaptive refining of time points at which new cohorts are introduced
into the population. In the original \textsc{EBT}, cohorts were introduced at a
fixed time interval. Under competition, the growth trajectories of
individuals born at similar times can diverge substantially over time
(Fig. \ref{fig:patch}a; Appendix \ref{sec:demography},
\ref{sec:patch} and \ref{sec:cohort-spacing}). This can become a
problem later during patch development because much of the population
can be located between widely-spaced cohorts, delivering low numerical
precision. To maintain numerical precision we therefore use an
iterative algorithm to adaptively refine the cohort spacing until the
trajectories of adjacent cohorts remain adequately resolved throughout
development of the patch (see Appendix \ref{sec:demography} for
details). This refinement causes an uneven distribution of cohort
introduction times across patch age (Fig.  \ref{fig:patch}a), with
considerably tighter spacing of cohorts in younger patches.

@@

Modelling of size-structured dynamics via solving of a deterministic \textsc{pde}
enables multiple complex non-linear demographic phenomena to be
investigated. First, {\plant} can simulate patterns of patch development after
disturbance, including self-thinning behaviours and successional
replacement. Whereas previous approaches for modelling self-thinning
were limited to a single species
\citep[e.g.][]{Barnes-2004, Coomes-2007}, {\plant} accommodates multiple
 species with different traits (Fig. \ref{fig:schematic}b; Fig.
\ref{fig:patch}).

Second, {\plant} allows for the effects of competition and succession on
productivity and biomass accumulation within a patch to be investigated
\citep{Falster-2011}. Leaf area cover and rates of biomass production
tend to vary with patch age. Such changes might come about via a
combination of physiological and structural changes
\citep{Binkley-2002, Smith-2001, Coomes-2007, Ogawa-2010}. {\plant} enables
the effects of competitive feedbacks to be incorporated, in addition to
any intrinsic physiological factors. Moreover, {\plant} allows for the
contributions of different species to be mapped (Fig. \ref{fig:patch}c).
